---
title: "RNAseq, multispecies killifish osmotic challenge"
author: "Lisa K. Johnson"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    df_print: paged
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float: yes
---

```{r LoadPackages, results='hide', include=FALSE}
# Install function for packages    
packages<-function(x){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x,repos="http://cran.r-project.org")
    require(x,character.only=TRUE)
  }
}
bioconductors <- function(x){
    x<- as.character(match.call()[[2]])
    if (!require(x, character.only = TRUE)){
      source("https://bioconductor.org/biocLite.R")
      biocLite(pkgs=x)
      require(x, character.only = TRUE)
    }
}
# for QC
packages("RColorBrewer")
packages("pheatmap")
packages("vsn")
packages(pheatmap)
# for DESeq

packages(lattice)
packages(RColorBrewer)
packages(dplyr)
packages(tidyr)
packages(ggplot2)
bioconductors(DESeq2)
bioconductors(tximport)
bioconductors(biomaRt)
```


```{r load custom scripts and data, results='hide', include=FALSE}
if(!file.exists('plotPCAWithSampleNames.R')){
  download.file('https://gist.githubusercontent.com/ljcohen/d6cf3367efae60d7fa1ea383fd7a1296/raw/f11030ced8c7953be6fada0325b92c20d369e0a7/plotPCAWithSampleNames.R', 'plotPCAWithSampleNames.R')
  }
source('plotPCAWithSampleNames.R')
if(!file.exists('overLapper_original.R')){
  download.file("https://gist.githubusercontent.com/ljcohen/b7e5fa93b8b77c33d26e4c44cadb5bb7/raw/3a2f84be2e937ef721cafaf308ff6456168f30d9/overLapper_original.R",'overLapper_original.R')
}
source('overLapper_original.R')
# This is just the counts with Experimental Design Info in the last 5 rows
if(!file.exists('../../Ensembl_species_counts_designfactors.csv')){
  download.file("https://osf.io/7vp38/download",'../../Ensembl_species_counts_designfactors.csv')
}
counts_design <- read.csv("../../Ensembl_species_counts_designfactors.csv",stringsAsFactors = FALSE)
```


```{r format counts and ExpDesign, include=FALSE, results='hide'}
design <- counts_design[counts_design$Ensembl == 'Empty',]
#design$type <- c("species","native_salinity","clade","group","condition")
drops <- c("X","Ensembl")
counts<-counts_design[!counts_design$Ensembl == 'Empty',]
rownames(counts)<-counts$Ensembl
design <- design[ , !(names(design) %in% drops)]
counts <- counts[ , !(names(counts) %in% drops)]
dim(design)
dim(counts)
```

```{r design for DESeq, results='hide', include=FALSE}
# design cateogories (full)
species<-as.character(unlist(design[1,]))
nativephysiology<-as.character(unlist(design[2,]))
clade<-as.character(unlist(design[3,]))
np_cl<-as.character(unlist(design[4,]))
condition<-as.character(unlist(design[5,]))
species_physiology<-as.vector(paste(np_cl, species, sep="_"))
species_condition<-as.vector(paste(species,condition,sep="_"))
species_group<-as.vector(paste(np_cl,species,sep="_"))
species_clade<-as.vector(paste(clade,species,sep="_"))
```
# PCA, all counts

This is the PCA with all counts, not filtered. The dimensions of the counts table are listed below. Row=genes, Columns=samples
```{r PCA of counts,}
# normal full counts
x <- data.matrix(counts)
dim(x)
x <- x+1
log_x<-log(x)
names<-colnames(log_x)
pca = prcomp(t(log_x))
summary(pca)
fac = factor(nativephysiology)
colours = function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))])}
mar.default <- c(5,4,4,2) + 0.1
par(mar = mar.default + c(0, 4, 0, 0)) 
plot(pca$x[,1:2], 
     col=colours(clade), 
     pch = c(16, 2, 9)[as.numeric(as.factor(nativephysiology))],
     cex=2,
     xlab="PC1",
     ylab="PC2",
     cex.lab=2,
     cex.axis = 2)
legend(140,100,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(clade))),cex=0.75, pch=19)
legend(140,-67,legend=c("Brackish","Freshwater","Marine"),cex=0.75,pch=c(16, 2, 9))
```

# Filtering counts

The following shows the dimensions of the dataframe when we filter out genes with low counts.

## At least 100 samples have a count of at least 10:

```{r filtering1,}
filter <- rownames(counts[rowSums(counts >= 10) >= 100,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

## At least 100 samples have a count of at least 1:

```{r filtering2,}
filter <- rownames(counts[rowSums(counts >= 1) >= 100,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

## At least 100 samples have a count of at least 0.1:

```{r filtering3,}
filter <- rownames(counts[rowSums(counts >= 0.1) >= 100,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```
## At least 100 samples have a count of at least 0.01:

```{r filtering4,}
filter <- rownames(counts[rowSums(counts >= 0.01) >= 100,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

PCA of at least 100 samples with a count of at least 0.01

```{r filtered4 counts PCA,}
# PCA
x <- data.matrix(filtered_counts)
x <- x+1
log_x<-log(x)
names<-colnames(log_x)
pca = prcomp(t(log_x))
summary(pca)
fac = factor(nativephysiology)
colours = function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))])}
mar.default <- c(5,4,4,2) + 0.1
par(mar = mar.default + c(0, 4, 0, 0)) 
plot(pca$x[,1:2], 
     col=colours(clade), 
     pch = c(16, 2, 9)[as.numeric(as.factor(nativephysiology))],
     cex=2,
     xlab="PC1",
     ylab="PC2",
     cex.lab=2,
     cex.axis = 2)
legend(-180,100,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(clade))),cex=0.75, pch=19)
legend(-180,-30,legend=c("Brackish","Freshwater","Marine"),cex=0.75,pch=c(16, 2, 9))
```

## All 128 samples must have a count of at least 10:

```{r filtering5,}
filter <- rownames(counts[rowSums(counts >= 10) >= 128,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

## All 128 samples must have a count of at least 1:

```{r filtering6,}
# filtering counts
filter <- rownames(counts[rowSums(counts >= 1) >= 128,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

## All 128 samples have a count of at least 0.1:

```{r filtering7,}
filter <- rownames(counts[rowSums(counts >= 0.1) >= 128,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

## All 128 samples have a count of at least 0.01:

```{r filtering8,}
filter <- rownames(counts[rowSums(counts >= 0.01) >= 128,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```

PCA of 128 samples with a count of at least 0.01:

```{r filtered8 counts PCA,}
# PCA
x <- data.matrix(filtered_counts)
x <- x+1
log_x<-log(x)
names<-colnames(log_x)
pca = prcomp(t(log_x))
summary(pca)
fac = factor(nativephysiology)
colours = function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))])}
mar.default <- c(5,4,4,2) + 0.1
par(mar = mar.default + c(0, 4, 0, 0)) 
plot(pca$x[,1:2], 
     col=colours(clade), 
     pch = c(16, 2, 9)[as.numeric(as.factor(nativephysiology))],
     cex=2,
     xlab="PC1",
     ylab="PC2",
     cex.lab=2,
     cex.axis = 2)
legend(-140,45,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(clade))),cex=0.75, pch=19)
legend(-140,-15,legend=c("Brackish","Freshwater","Marine"),cex=0.75,pch=c(16, 2, 9))
```

# Contrasts

```{r PCA of log2FC contrasts,results='hide', include=FALSE}
if(!file.exists('../../BW_FW_contrasts_design.csv')){
  download.file("https://osf.io/wvqxs/download",'../../BW_FW_contrasts_design.csv')
}
if(!file.exists('../../TR_BW_contrasts_design.csv')){
  download.file("https://osf.io/vwj2n/download",'../../TR_BW_contrasts_design.csv')
}
if(!file.exists('../../TR_FW_contrasts_design.csv')){
  download.file("https://osf.io/59en7/download",'../../TR_FW_contrasts_design.csv')
}
contrast_BW_v_FW <- read.csv("../../BW_FW_contrasts_design.csv",stringsAsFactors = FALSE)
contrast_TR_v_BW <- read.csv("../../TR_BW_contrasts_design.csv",stringsAsFactors = FALSE)
contrast_TR_v_FW <- read.csv("../../TR_FW_contrasts_design.csv",stringsAsFactors = FALSE)
TR_v_FW_design <- contrast_TR_v_FW[c(29248:29251),]  
BW_v_FW_design <- contrast_BW_v_FW[c(29248:29251),]
TR_v_BW_design <- contrast_TR_v_BW[c(29248:29251),]
contrast_TR_v_FW <- contrast_TR_v_FW[-c(29248:29251),]
contrast_BW_v_FW <- contrast_BW_v_FW[-c(29248:29251),]
contrast_TR_v_BW <- contrast_TR_v_BW[-c(29248:29251),]
# design factors
de<-as.character(unlist(TR_v_FW_design[4,]))
de<-de[-c(1)]
cl<-as.character(unlist(TR_v_FW_design[3,]))
cl<-cl[-c(1)]
ph<-as.character(unlist(TR_v_FW_design[2,]))
ph<-ph[-c(1)]
names<-contrast_TR_v_FW$X
rownames(contrast_TR_v_FW) <- names
head(contrast_TR_v_FW)
dim(contrast_TR_v_FW)
drops <- c("X")
contrast_TR_v_FW <- contrast_TR_v_FW[ , !(names(contrast_TR_v_FW) %in% drops)]
contrast_TR_v_FW <- na.omit(contrast_TR_v_FW)
dim(contrast_TR_v_FW)
df_contrast_TR_v_FW<-data.matrix(contrast_TR_v_FW)
# get rownames where counts are at least 1 in all 128 samples
filter <- rownames(counts[rowSums(counts >= 1) >= 128,])
filtered_counts <- counts[filter,]
filtered_names <- rownames(filtered_counts)
filtered_contrast_TR_v_FW <- contrast_TR_v_FW[rownames(contrast_TR_v_FW) %in% filtered_names,]
dim(filtered_contrast_TR_v_FW)
x <- filtered_contrast_TR_v_FW
dim(x)
x <- na.omit(x)
dim(x)
x <- data.matrix(x)
colnames(x)
names<-colnames(x)
pca = prcomp(t(x))
fac = factor(ph)
fac
colours = function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))])}
summary(pca)
mar.default <- c(5,4,4,2) + 0.1
par(mar = mar.default + c(0, 4, 0, 0)) 
plot(pca$x[,1:2], 
     col=colours(cl), 
     #pch=19,
     pch = c(16, 2, 9)[as.numeric(as.factor(ph))],
     cex=2,
     xlab="PC1",
     ylab="PC2",
     cex.lab=2,
     cex.axis = 2)
legend(-40,30,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(cl))),cex=1.5, pch=19)
legend(-40,10,legend=c("Brackish native","Freshwater native","Marine native"),cex=1.5,pch=c(16, 2, 9))
#text(pca$x[,1:2], labels=names, pos=3)
dev.off()
```

# DESeq2 analysis

## Design

```{r DESeq2 design,}
cols<-colnames(filtered_counts)
ExpDesign <- data.frame(row.names=cols, condition=condition, species = species, clade=clade,species_condition=species_condition,species_clade = species_clade)
ExpDesign
```

Clade1_F_catanatus and 0.2 ppt are reference levels for comparison.

```
m1 <- model.matrix(~condition + species_clade + species_clade:condition,ExpDesign)
```

```{r run DESeq, results='hide', include=FALSE}
m1 <- model.matrix(~condition + species_clade + species_clade:condition,ExpDesign)
colnames(m1)
all.zero <- apply(m1, 2, function(x) all(x==0))
idx <- which(all.zero)
m1 <- m1[,-idx]
all(rownames(ExpDesign) == colnames(filtered_counts))
counts_round<- round(data.matrix(filtered_counts),digits=0)
dds <- DESeqDataSetFromMatrix(countData = counts_round,colData = ExpDesign,design = m1)
dds <- DESeq(dds, full = m1, betaPrior=FALSE)
ddsClean <- dds[which(mcols(dds)$betaConv),]
dds<-ddsClean
```

```{r QA of DESeq,}
resultsNames(dds)
vsd <- vst(dds, blind=FALSE)
meanSdPlot(assay(vsd))
plotDispEsts(dds)
```

```{r PCA of DESeq,}
plotPCA(vsd, intgroup=c("species_clade"))
plotPCAWithSampleNames(vsd,intgroup=c("species_clade"))
```

```{r Display versions,}
sessionInfo()
```
