
---
title: "RNAseq, DE analysis for F_rathbuni killifish osmotic challenge"
author: "Lisa K. Johnson"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    df_print: paged
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float: yes
---


```{r LoadPackages, results='hide', include=FALSE}
packages<-function(x){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x,repos="http://cran.r-project.org")
    require(x,character.only=TRUE)
  }
}
bioconductors <- function(x){
    x<- as.character(match.call()[[2]])
    if (!require(x, character.only = TRUE)){
      source("https://bioconductor.org/biocLite.R")
      biocLite(pkgs=x)
      require(x, character.only = TRUE)
    }
}
packages("RColorBrewer")
packages("pheatmap")
packages("vsn")
packages(pheatmap)
packages(cowplot)
# for DESeq
packages(lattice)
packages(RColorBrewer)
packages(dplyr)
packages(tidyr)
packages(gplots)
packages(ggplot2)
bioconductors(DESeq2)
bioconductors(tximport)
bioconductors(biomaRt)
```


```{r load custom scripts and data, results='hide', include=FALSE}
if(!file.exists('plotPCAWithSampleNames.R')){
  download.file('https://gist.githubusercontent.com/ljcohen/d6cf3367efae60d7fa1ea383fd7a1296/raw/f11030ced8c7953be6fada0325b92c20d369e0a7/plotPCAWithSampleNames.R', 'plotPCAWithSampleNames.R')
	}
source('plotPCAWithSampleNames.R')
if(!file.exists('overLapper_original.R')){
  download.file("https://gist.githubusercontent.com/ljcohen/b7e5fa93b8b77c33d26e4c44cadb5bb7/raw/3a2f84be2e937ef721cafaf308ff6456168f30d9/overLapper_original.R",'overLapper_original.R')
	}
source('overLapper_original.R')
# This is just the counts with Experimental Design Info in the last 5 rows
if(!file.exists('../../Ensembl_species_counts_designfactors.csv')){
  download.file("https://osf.io/7vp38/download",'../../Ensembl_species_counts_designfactors.csv')
	}
counts_design <- read.csv("../../Ensembl_species_counts_designfactors.csv",stringsAsFactors = FALSE)
```


```{r format counts and ExpDesign,}
design <- counts_design[counts_design$Ensembl == 'Empty',]
drops <- c("X","Ensembl")
counts<-counts_design[!counts_design$Ensembl == 'Empty',]
rownames(counts)<-counts$Ensembl
design <- design[ , !(names(design) %in% drops)]
counts <- counts[ , !(names(counts) %in% drops)]
design <- design[ , startsWith(names(design),"F_rathbuni")]
counts <- counts[ , startsWith(names(counts),"F_rathbuni")]
dim(design)
dim(counts)
# design cateogories (full)
species<-as.character(unlist(design[1,]))
nativephysiology<-as.character(unlist(design[2,]))
clade<-as.character(unlist(design[3,]))
condition<-as.character(unlist(design[5,]))
cols<-colnames(counts)
ExpDesign <- data.frame(row.names=cols,
                        condition=condition)
ExpDesign
```


# Filtering counts

The following shows the dimensions of the dataframe when we filter out genes with low counts.

2 samples must have a count of at least 0.1:

```{r filtering5,}
filter <- rownames(counts[rowSums(counts >= 2) >= 0.1,])
filtered_counts <- counts[filter,]
dim(filtered_counts)
```


```{r run DESeq, results='hide', include=FALSE}
all(rownames(ExpDesign) == colnames(counts))
counts_round<- round(data.matrix(counts),digits=0)
dds <- DESeqDataSetFromMatrix(countData = counts_round,colData = ExpDesign,design = ~condition)
dds<-DESeq(dds)
```


# DESeq

```{r DESeq QC, }
plotDispEsts(dds)
resultsNames(dds)
vsd <- vst(dds, blind=FALSE)
meanSdPlot(assay(vsd))
```



# PCA

```{r PCA of counts,}
plotPCA(vsd, intgroup=c("condition"))
plotPCAWithSampleNames(vsd,intgroup=c("condition"))
```


# MA plot, 15_ppt vs. 0.2_ppt
```{r MA plot 1, }
gene_id <- c("avpr2aa","slc24a5","CLDN4","aqp3","cftr","kcnj2a","polyamine-modulated factor 1-like","kcnj1a.6","sept2B","septin-2", "cipcb","clcn2c","zymogen granule membrane protein 16","atp1a1b","solute carrier family 24 member 2")
protein_id <- c("ENSFHEP00000000036","ENSFHEP00000001609","ENSFHEP00000003908","ENSFHEP00000006725","ENSFHEP00000008393","ENSFHEP00000009753","ENSFHEP00000013324",
                "ENSFHEP00000015383","ENSFHEP00000015765","ENSFHEP00000016853","ENSFHEP00000017303","ENSFHEP00000019510","ENSFHEP00000025841",
                "ENSFHEP00000031108","ENSFHEP00000034177")
res<-results(dds,contrast=c("condition","15_ppt","0.2_ppt"))
res_ordered <-as.data.frame(res[order(res$padj),])
res_filtered <-subset(res_ordered,res_ordered$padj<0.05)
id<-rownames(res_filtered)
res_filtered<-cbind(res_filtered,id)
plot(log2(res$baseMean), res$log2FoldChange, 
     col=ifelse(res$padj < 0.05, "red","gray67"),
     main="F_rathbuni (15_ppt vs. 0.2_ppt) (padj<0.05)",xlim=c(1,15),pch=20,cex=1)
abline(h=c(-1,1), col="blue")
resSig = res_ordered[rownames(res_ordered) %in% protein_id, ]
dim(resSig)
genes<-rownames(resSig)
mygenes <- resSig[,]
baseMean_mygenes <- mygenes[,"baseMean"]
log2FoldChange_mygenes <- mygenes[,"log2FoldChange"]
text(log2(baseMean_mygenes),log2FoldChange_mygenes,labels=gene_id,pos=2,cex=0.60)
```


# MA plot, transfer vs. 0.2_ppt
```{r MA plot 2, }
gene_id <- c("avpr2aa","slc24a5","CLDN4","aqp3","cftr","kcnj2a","polyamine-modulated factor 1-like","kcnj1a.6","sept2B","septin-2", "cipcb","clcn2c","zymogen granule membrane protein 16","atp1a1b","solute carrier family 24 member 2")
protein_id <- c("ENSFHEP00000000036","ENSFHEP00000001609","ENSFHEP00000003908","ENSFHEP00000006725","ENSFHEP00000008393","ENSFHEP00000009753","ENSFHEP00000013324",
                "ENSFHEP00000015383","ENSFHEP00000015765","ENSFHEP00000016853","ENSFHEP00000017303","ENSFHEP00000019510","ENSFHEP00000025841",
                "ENSFHEP00000031108","ENSFHEP00000034177")
res<-results(dds,contrast=c("condition","transfer","0.2_ppt"))
res_ordered <-as.data.frame(res[order(res$padj),])
res_filtered <-subset(res_ordered,res_ordered$padj<0.05)
id<-rownames(res_filtered)
res_filtered<-cbind(res_filtered,id)
plot(log2(res$baseMean), res$log2FoldChange, 
     col=ifelse(res$padj < 0.05, "red","gray67"),
     main="F_rathbuni (transfer vs. 0.2_ppt) (padj<0.05)",xlim=c(1,15),pch=20,cex=1)
abline(h=c(-1,1), col="blue")
resSig = res_ordered[rownames(res_ordered) %in% protein_id, ]
dim(resSig)
genes<-rownames(resSig)
mygenes <- resSig[,]
baseMean_mygenes <- mygenes[,"baseMean"]
log2FoldChange_mygenes <- mygenes[,"log2FoldChange"]
text(log2(baseMean_mygenes),log2FoldChange_mygenes,labels=gene_id,pos=2,cex=0.60)
```


# MA plot, transfer vs. 15_ppt
```{r MA plot 3, }
gene_id <- c("avpr2aa","slc24a5","CLDN4","aqp3","cftr","kcnj2a","polyamine-modulated factor 1-like","kcnj1a.6","sept2B","septin-2", "cipcb","clcn2c","zymogen granule membrane protein 16","atp1a1b","solute carrier family 24 member 2")
protein_id <- c("ENSFHEP00000000036","ENSFHEP00000001609","ENSFHEP00000003908","ENSFHEP00000006725","ENSFHEP00000008393","ENSFHEP00000009753","ENSFHEP00000013324",
                "ENSFHEP00000015383","ENSFHEP00000015765","ENSFHEP00000016853","ENSFHEP00000017303","ENSFHEP00000019510","ENSFHEP00000025841",
                "ENSFHEP00000031108","ENSFHEP00000034177")
res<-results(dds,contrast=c("condition","transfer","15_ppt"))
res_ordered <-as.data.frame(res[order(res$padj),])
res_filtered <-subset(res_ordered,res_ordered$padj<0.05)
id<-rownames(res_filtered)
res_filtered<-cbind(res_filtered,id)
plot(log2(res$baseMean), res$log2FoldChange, 
     col=ifelse(res$padj < 0.05, "red","gray67"),
     main="F_rathbuni (transfer vs. 15_ppt) (padj<0.05)",xlim=c(1,15),pch=20,cex=1)
abline(h=c(-1,1), col="blue")
resSig = res_ordered[rownames(res_ordered) %in% protein_id, ]
dim(resSig)
genes<-rownames(resSig)
mygenes <- resSig[,]
baseMean_mygenes <- mygenes[,"baseMean"]
log2FoldChange_mygenes <- mygenes[,"log2FoldChange"]
text(log2(baseMean_mygenes),log2FoldChange_mygenes,labels=gene_id,pos=2,cex=0.60)
```

# Salinity-responsive genes of interest

## avpr2aa
```{r plot goi 1,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000000036"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("avpr2aa")
plot(C1)
```


## slc24a5
```{r plot goi 2,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000001609"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("slc24a5")
plot(C1)
```


## CLDN4
```{r plot goi 3,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000003908"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("CLDN4")
plot(C1)
```


## aqp3
```{r plot goi 4,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000006725"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("aqp3")
plot(C1)
```


## cftr
```{r plot goi 5,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000008393"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("cftr")
plot(C1)
```


## kcnj2a
```{r plot goi 6,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000009753"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("kcnj2a")
plot(C1)
```


## polyamine-modulated factor 1-like
```{r plot goi 7,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000013324"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("polyamine-modulated factor 1-like")
plot(C1)
```


## kcnj1a.6
```{r plot goi 8,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000015383"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("kcnj1a.6")
plot(C1)
```


## sept2B
```{r plot goi 9,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000015765"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("sept2B")
plot(C1)
```


## septin-2
```{r plot goi 10,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000016853"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("septin-2")
plot(C1)
```


## cipcb
```{r plot goi 11,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000017303"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("cipcb")
plot(C1)
```


## clcn2c
```{r plot goi 12,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000019510"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("clcn2c")
plot(C1)
```


## zymogen granule membrane protein 16
```{r plot goi 13,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000025841"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("zymogen granule membrane protein 16")
plot(C1)
```


## atp1a1b
```{r plot goi 14,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000031108"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("atp1a1b")
plot(C1)
```


## solute carrier family 24 member 2
```{r plot goi 15,}
tcounts <- t(log2((counts(dds[c("ENSFHEP00000034177"), ], normalized=TRUE, replaced=FALSE)+.5))) %>% 
  merge(colData(dds), ., by="row.names") %>% 
  gather(gene, expression, (ncol(.)-1+1):ncol(.))

C1<-ggplot(tcounts, aes(condition, expression)) +
  geom_point() + 
  stat_summary(fun.y="mean", geom="line") +
  stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
               geom="errorbar",width=0.2) +
  theme_bw() +
  theme(legend.position="none",panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y="Expression (log2 normalized counts)")+
  ggtitle("solute carrier family 24 member 2")
plot(C1)
```



# Get normalized counts and filter out genes with low expression

This is the number of genes in the expression counts table:
```{r dim counts, }
# get counts
counts_table = counts(dds, normalized=TRUE)
dim(counts_table)
```

After filtering for low expression (where rowSum is greater than or equal to 1):

```{r filtering again, }
filtered_norm_counts<-counts_table[!rowSums(counts_table==0)>=1, ]
dim(filtered_norm_counts)
filtered_norm_counts<-as.data.frame(filtered_norm_counts)
GeneID<-rownames(filtered_norm_counts)
filtered_norm_counts<-cbind(filtered_norm_counts,GeneID)
```


```{r biomaRt,}
all_goi<-c("ENSFHEP00000007220.1","ENSFHEP00000025841","ENSFHEP00000019510",
           "ENSFHEP00000015383","ENSFHEP00000009753","ENSFHEP00000006725","ENSFHEP00000008393",
           "ENSFHEP00000013324","ENSFHEP00000001609","ENSFHEP00000013324","ENSFHEP00000034177",
           "ENSFHEP00000015765","ENSFHEP00000017303","ENSFHEP00000000036","ENSFHEP00000031108",
           "ENSFHEP00000016853","ENSFHEP00000003908")

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
ensembl = useDataset("fheteroclitus_gene_ensembl",mart=ensembl)
ensembl_proteinID = rownames(counts_table)
query<-getBM(attributes=c('ensembl_peptide_id','ensembl_transcript_id',
'ensembl_gene_id','gene_biotype','external_gene_name',
'description','entrezgene'), filters = 'ensembl_peptide_id', values = ensembl_proteinID, mart=ensembl)

# link goi Ensembl ID to external_gene_name or description
gene_id <- c("avpr2aa","slc24a5","CLDN4","aqp3","cftr","kcnj2a","polyamine-modulated factor 1-like","kcnj1a.6","sept2B","septin-2", "cipcb","clcn2c","zymogen granule membrane protein 16","atp1a1b","solute carrier family 24 member 2")
protein_id <- c("ENSFHEP00000000036","ENSFHEP00000001609","ENSFHEP00000003908","ENSFHEP00000006725","ENSFHEP00000008393","ENSFHEP00000009753","ENSFHEP00000013324",
                "ENSFHEP00000015383","ENSFHEP00000015765","ENSFHEP00000016853","ENSFHEP00000017303","ENSFHEP00000019510","ENSFHEP00000025841",
                "ENSFHEP00000031108","ENSFHEP00000034177")
```


```{r merge and write, }
# -----------------------------
# stats results
# -----------------------------

res_BW_v_FW <- results(dds, tidy=TRUE, contrast=c("condition","15_ppt","0.2_ppt")) %>% arrange(padj) %>% tbl_df() 
res_TR_v_FW <- results(dds, tidy=TRUE, contrast=c("condition","transfer","0.2_ppt")) %>% arrange(padj) %>% tbl_df() 
res_TR_v_BW <- results(dds, tidy=TRUE, contrast=c("condition","transfer","15_ppt")) %>% arrange(padj) %>% tbl_df() 

# -----------------------------
# counts 
# -----------------------------

cols <- colnames(counts_table)
counts_table <- as.data.frame(counts_table[,cols])
dim(counts_table)

# -----------------------------
# column names for stats from BW_v_FW specific contrast
# -----------------------------

names(res_BW_v_FW)[names(res_BW_v_FW) == 'padj'] <- 'padj-15ppt-v-0.2ppt'
names(res_BW_v_FW)[names(res_BW_v_FW) == 'baseMean'] <- 'baseMean-ALL'
names(res_BW_v_FW)[names(res_BW_v_FW) == 'log2FoldChange'] <- 'log2FoldChange-15ppt-v-0.2ppt'
names(res_BW_v_FW)[names(res_BW_v_FW) == 'lfcSE'] <- 'lfcSE-15ppt-v-0.2ppt'
names(res_BW_v_FW)[names(res_BW_v_FW) == 'stat'] <- 'stat-15ppt-v-0.2ppt'
names(res_BW_v_FW)[names(res_BW_v_FW) == 'pvalue'] <- 'pvalue-15ppt-v-0.2ppt'

# -----------------------------
# column names for stats from TR_v_FW specific contrast
# -----------------------------

names(res_TR_v_FW)[names(res_TR_v_FW) == 'padj'] <- 'padj-TR-v-0.2ppt'
names(res_TR_v_FW)[names(res_TR_v_FW) == 'baseMean'] <- 'baseMean-ALL'
names(res_TR_v_FW)[names(res_TR_v_FW) == 'log2FoldChange'] <- 'log2FoldChange-TR-v-0.2ppt'
names(res_TR_v_FW)[names(res_TR_v_FW) == 'lfcSE'] <- 'lfcSE-TR-v-0.2ppt'
names(res_TR_v_FW)[names(res_TR_v_FW) == 'stat'] <- 'stat-TR-v-0.2ppt'
names(res_TR_v_FW)[names(res_TR_v_FW) == 'pvalue'] <- 'pvalue-TR-v-0.2ppt'

# -----------------------------
# column names for stats from TR_v_BW specific contrast
# -----------------------------

names(res_TR_v_BW)[names(res_TR_v_BW) == 'padj'] <- 'padj-TR-v-15ppt'
names(res_TR_v_BW)[names(res_TR_v_BW) == 'baseMean'] <- 'baseMean-ALL'
names(res_TR_v_BW)[names(res_TR_v_BW) == 'log2FoldChange'] <- 'log2FoldChange-TR-v-15ppt'
names(res_TR_v_BW)[names(res_TR_v_BW) == 'lfcSE'] <- 'lfcSE-TR-v-15ppt'
names(res_TR_v_BW)[names(res_TR_v_BW) == 'stat'] <- 'stat-TR-v-15ppt'
names(res_TR_v_BW)[names(res_TR_v_BW) == 'pvalue'] <- 'pvalue-TR-v-15ppt'

# -----------------------------
# merge counts and stats
# -----------------------------
res_TR_v_FW <- as.data.frame(res_TR_v_FW)
rownames(res_TR_v_FW) <- res_TR_v_FW$row
counts_table_stats <- merge(as.data.frame(res_TR_v_FW),counts_table,by=0)
counts_table_stats <- merge(as.data.frame(res_BW_v_FW),counts_table_stats,by='row')
counts_table_stats <- merge(as.data.frame(res_TR_v_BW),counts_table_stats,by='row')
dim(counts_table_stats)

# -----------------------------
# merge annotations with stats
# -----------------------------
counts_table_ann <- merge(query,counts_table_stats,by.x = "ensembl_peptide_id", by.y = "row", all = TRUE)
counts_table_ann <- counts_table_ann[!duplicated(counts_table_ann$ensembl_peptide_id), ]
rownames(counts_table_ann) <- counts_table_ann$ensembl_peptide_id
dim(counts_table_ann)
counts_table_ann <- counts_table_ann[ , -which(names(counts_table_ann) %in% c("Row.names"))]
# -----------------------------
# write csv files
# -----------------------------
counts_table_ann <- counts_table_ann[order(counts_table_ann[,19]), ]
write.csv(counts_table_ann,"/Users/johnsolk/Documents/UCDavis/Whitehead/counts_stats_byspecies/F_rathbuni_stats_annotations_counts.csv")
```

